Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-26 paper/paper.tex
floor notation
finally proved iso
Add a paper-overview picture
intro, contributions
Remove hyphen in 'algebraically-indexed' everywhere

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Abstraction and Invariance for Algebraically-Indexed Types,
            pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

@@

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}

@@

\newcommand{\relEnv}[1]{\mathcal{#1}}
%\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
%\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
%\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
%\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
\newcommand{\tySem}[1]{\left| #1\right|}
\newcommand{\ctxtSem}[1]{\left| #1\right|}
\newcommand{\tmSem}[1]{\left| #1\right|}
\newcommand{\tyPrimSem}[1]{\left| #1\right|}
\newcommand{\rsem}[1]{\llbracket #1 \rrbracket}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

@@

\bibitem{kennedy97relational} 
A.~Kennedy. Relational Parametricity and Units of Measure.
Proceedings, POPL, pp. 442-455, 1997.

>>>>

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Abstraction and Invariance for Algebraically Indexed Types,
            pdfauthor={Robert Atkey, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

@@

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{pgf}
\usepackage{tikz}
\usepgflibrary{shapes.geometric}
\usepgflibrary{shapes.misc}
\usepgflibrary{shapes.multipart}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.multipart}

@@

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\ctxtSem}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\tmSem}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\tyPrimSem}[1]{\left\lfloor #1 \right\rfloor}
%\newcommand{\tySem}[1]{\left| #1\right|}
%\newcommand{\ctxtSem}[1]{\left| #1\right|}
%\newcommand{\tmSem}[1]{\left| #1\right|}
%\newcommand{\tyPrimSem}[1]{\left| #1\right|}
\newcommand{\rsem}[1]{\llbracket #1 \rrbracket}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

@@

\bibitem{kennedy97relational} 
A.~J.~Kennedy. Relational Parametricity and Units of Measure.
Proceedings, POPL, pp. 442-455, 1997.

\bibitem{fsharp}
A.~J.~Kennedy. Types for Units-of-Measure: Theory and Practice.
Central European Functional Programming school (CEFP), pp. 268--305, LNCS vol. 6299, 2010.

### 2012-10-26 paper/section-general-framework.tex
Remove hyphen in 'algebraically-indexed' everywhere

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Index-Erasure Semantics}
\label{sec:erasure-semantics}
%\fixme{Make sure that the semantics is motivated}
Having defined the syntax of algebraically-indexed types and terms,
we turn to their denotational interpretation.  We first define an
\emph{index-erasure} interpretation of types and terms that interprets
every well-indexed type as a set, ignoring the indexing
expressions, and which interprets open terms as functions that map environments
to final values.
%Building on the index-erasure semantics, we
%We then define in
%\autoref{sec:relational-semantics} the relational
%interpretation of types.

@@

An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
interpreted as cartesian product, i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
\times\cdots \times \semSort{s_n}$. 
For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of~$e$:
\begin{align*}
\semIndexExp{i} \rho & \isDefinedAs \rho(i) &
\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)} \rho & \isDefinedAs \semIndexExp{\indexOp{f}}(
\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)
\end{align*}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Index-Erasure Semantics}
\label{sec:erasure-semantics}
%\fixme{Make sure that the semantics is motivated}
Having defined the syntax of algebraically indexed types and terms,
we turn to their denotational interpretation.  We first define an
\emph{index-erasure} interpretation of types and terms that interprets
every well-indexed type as a set, ignoring the indexing
expressions, and which interprets open terms as functions that map environments
to final values.
%Building on the index-erasure semantics, we
%We then define in
%\autoref{sec:relational-semantics} the relational
%interpretation of types.

@@

An index context $\Delta = (i_1\mathord:s_1,\ldots,i_n\mathord:s_n)$ is
interpreted as cartesian product,x i.e., $\semSort{\Delta} \isDefinedAs \semSort{s_1}
\times\cdots \times \semSort{s_n}$. 
For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of~$e$:
\begin{align*}
\semIndexExp{i} \rho & \isDefinedAs \rho(i) &
\semIndexExp{\indexOp{f}(e_1, \ldots, e_n)} \rho & \isDefinedAs \semIndexExp{\indexOp{f}}(
\semIndexExp{e_1}\rho,\ldots,\semIndexExp{e_n}\rho)
\end{align*}

### 2012-10-26 paper/section-introduction.tex
Remove hyphen in 'algebraically-indexed' everywhere

Our motivations for studying algebraically-indexed types are
threefold. First, we believe that, as with
units-of-measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

We follow the mantra \emph{semantics first, syntax later} in studying
types with algebraic structure. We have not yet built a practical
programming language that supports algebraically-indexed types; nor
have we designed type checking, type inference, or static analysis
algorithms.  But when we do so, the semantics will guide us. The fact
that zero is polymorphic in units-of-measure (it can be given type $\forall
u.\tyPrim{real}u$) whereas other constants are dimensionless (having
type $\tyPrim{real}1$) is justified by the invariance properties
induced by the types: zero is invariant under scaling, other constants
are not. For less trivial constants and operations, the appropriate types are
not so apparent, as we shall see, but invariance properties expressed by
the semantics guide us in assigning appropriate types. Semantics does not lie!

>>>>

Our motivations for studying algebraically indexed types are
threefold. First, we believe that, as with
units-of-measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

We follow the mantra \emph{semantics first, syntax later} in studying
types with algebraic structure. We have not yet built a practical
programming language that supports algebraically indexed types; nor
have we designed type checking, type inference, or static analysis
algorithms.  But when we do so, the semantics will guide us. The fact
that zero is polymorphic in units-of-measure (it can be given type $\forall
u.\tyPrim{real}u$) whereas other constants are dimensionless (having
type $\tyPrim{real}1$) is justified by the invariance properties
induced by the types: zero is invariant under scaling, other constants
are not. For less trivial constants and operations, the appropriate types are
not so apparent, as we shall see, but invariance properties expressed by
the semantics guide us in assigning appropriate types. Semantics does not lie!

### 2012-10-26 paper/section-metric.tex
Refine statements of \eta_ops self-related-ness lemmas in Sections 6 and 7
Remove hyphen in 'algebraically-indexed' everywhere

For the relational intepretation we construct a model of the indexing
theory by interpreting $\metricSort$ with strictly positive real
numbers. We set $\rsem{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar
\abs{x - x'} < \epsilon \}$.
\begin{lemma}\label{lem:metric-environments}
  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\Metric,
  \eta_\Metric) \in \rsem{\Gamma}{\rho}$.
\end{lemma}

@@

\paragraph{Function Sensitivity}
A type system with a relational interpretation that tracks distances
between values has %previously
been investigated by Reed and Pierce \cite{reed10distance} in the
setting of differential privacy. Their system uses a linear type
discipline to ensure that all programs are $c$-sensitive (i.e., the
distance between the outputs is no greater than $c$ times the
distance between the inputs, for some constant $c$). We can express
their central concept of $c$-sensitivity (for functions on the reals)
as an algebraically-indexed type: $\forall
\epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to
\tyPrim{real}{\frac{1}{c}\epsilon}$. Investigating the precise
connection between their system and ours is left to future work.

>>>>

For the relational intepretation we construct a model of the indexing
theory by interpreting $\metricSort$ with strictly positive real
numbers. We set $\rsem{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar
\abs{x - x'} < \epsilon \}$.
\begin{lemma}\label{lem:metric-environments}
  $(\eta_\Metric, \eta_\Metric) \in \rsem{\Gamma_\Metric}{}*$.
\end{lemma}

@@

\paragraph{Function Sensitivity}
A type system with a relational interpretation that tracks distances
between values has %previously
been investigated by Reed and Pierce \cite{reed10distance} in the
setting of differential privacy. Their system uses a linear type
discipline to ensure that all programs are $c$-sensitive (i.e., the
distance between the outputs is no greater than $c$ times the
distance between the inputs, for some constant $c$). We can express
their central concept of $c$-sensitivity (for functions on the reals)
as an algebraically indexed type: $\forall
\epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to
\tyPrim{real}{\frac{1}{c}\epsilon}$. Investigating the precise
connection between their system and ours is left to future work.
